\section{Bernouilli's principle}

Let us approach Eq. \ref{eq:Euler_momentum_w_vort} in a completely
different manner. Instead of getting rid of the gradients, let us
express as many terms as we can as gradients, for reasons that will
become clear soon. First, the gravitational acceleration is, by
definition, $\bfg = -\nabla\varphi$, where $\varphi $ is the
gravitational potential energy ($g z$ if the $z$ axis is the
vertical). Now, the pressure gradient over the density may be related
to the gradient of pressure over density, thus:
\[
\nabla
\left(
\frac{p}{\rho}
\right) =
\frac{ \nabla p }{\rho} -
\frac{ \nabla p }{\rho^2} \nabla \rho 
\]
Therefore,
\[
\frac{\partial \bfu }{\partial t} +
 \nabla
\left(
\frac12 u^2 +
\frac{p}{\rho} +
\varphi
\right)
=
  \bfu \times\vort +
 \frac{ \nabla p }{\rho^2} \nabla \rho 
\]

Now, in steady flow we would have
\[
\nabla
\left(
\frac12 u^2 +
\frac{p}{\rho} +
\varphi
\right)
=
  \bfu \times\vort +
 \frac{ \nabla p }{\rho^2} \nabla \rho  .
\]

The term involving the vector product of the vorticity and the
velocity is perpendicular to both. Hence, it vanishes upon a scalar
multiplication with $\bfu$:
\[
\bfu\cdot
\nabla
\left(
\frac12 u^2 +
\frac{p}{\rho} +
\varphi
\right)
=
\frac{ \nabla p }{\rho^2} \bfu\cdot \nabla \rho  .
\]
 
In steady flow, the continuity equation is
\[
\bfu\cdot \nabla \rho  +
\rho \nabla \cdot \bfu = 0 ,
\]
hence if the flow is incompressible (technically, divergence-free, as
explained in section \ref{sec:}), $\bfu\cdot \nabla \rho $. I.e. the
density does not change along streamlines. Then,
\[
\bfu\cdot
\nabla
\left(
\frac12 u^2 +
\frac{p}{\rho} +
\varphi
\right)
=
0 ,
\]
which states the fact that the quantity
\[
h = \frac12 u^2 + \frac{p}{\rho} + \varphi
\]
is constant along a given streamline. This result is known as
Bernoulli's principle, and applies only to ideal, steady,
incompressible flow. (There is a variant of it that applies to
unsteady flow, as we will see in section \ref{sec:}.)
%
This combination is called ``the head'' and is customarily used in
elementary applications of this result. Some of its direct
applications are: the Venturi effect (by which the pressure decreases
in zones with higher velocities), slow drainage of containers, syphons
\ldots

We will also see that in some cases, like in potential flow, the
velocity field may be found independently of the pressure. This
principle then yields the corresponding pressure from the velocity.

%any field satisfies
%\[
%\frac{\partial \bfu }{\partial t} +
%\]

\section{Exercises}

\begin{enumerate}
\item \label{ex:vector_identity}) Prove that
\[
\bfu \times ( \nabla\times \bfq) =
%\frac12
\nabla_\bfq (\bfu\cdot\bfq) - \bfu \cdot \nabla \bfq ,
\]
for any two vector fields. By $\nabla_\bfq $ it is meant that the
gradient is applied only on $\bfq$ (Feyman's subscript notation).
Hint: the curl of a field is associated with a tensor with elements:
\[
\overline{( \nabla\times\bfq )_{ij}} =
\left(
\frac{\partial q_j}{\partial x_i} -
\frac{\partial q_i}{\partial x_j}
\right) \qquad
\overline{( \nabla\times\bfq )_{ij}} =
\sum_k \epsilon_{ijk}  ( \nabla\times\bfq )_k .
\]
In the latter, $\epsilon$ is a rank-three tensor, the totally
antisymmetric tensor (also known as the Levi-Civita symbol).  If this
seems mistifying, all that it is expressing is that e.g.  $
(\nabla\times\bfq)_z= \partial q_x /\partial y - \partial q_y
/\partial x$, et c\ae tera.

This tensor appears also in the vector product:
\[
(\bfu\times\bfq)_i= \sum_{j,k} \epsilon_{ijk} u_j q_k .
\]
These two results make the demonstration quite simple. We will
also be using these expressions in section \ref{sec:}, when
we talk about rotations of a particle.

\end{enumerate}
