\chapter*{}

\vspace{6ex}
%\begin{quotation}
\begin{quote}
  \greektext

  Ποταμο˜ισι το˜ισιν αὐτο˜ισιν ἐμβαίνουσιν, ἕτερα καὶ ἕτερα ὕδατα ἐπιρρε˜ι  \\%[0.5cm]
  \latintext
  \begin{em}
    Ever-newer waters flow on those who step into the same rivers.  \\%[1cm]
  \end{em}
  Heraclitus \cite{Diels-Kranz}.%, Fragmente der Vorsokratiker, 22 B12
\end{quote}

%\end{quotation}

\newpage

\chapter*{Foreword}


Just a compilation of notes on different aspects of fluid
dynamics I have collected over the years.

The final push was from teaching the course Fluid Dynamics on the
International Master of Nuclear Fusion, course 2018. I realized much
material was scattered all over the place, on notes, articles, blog
entries \ldots

I may upload these notes to a collaborative site (e.g. github) so that
other people may contribute. As of today, I am the sole author.\\[4cm]

Daniel Duque \\
CEHINAV (Naval Model Basin Research Group) \\
Universidad Polit\'ecnica de Madrid \\
Madrid, Spain 2018


\chapter{Introduction}

Fluids, by definition, flow.

The main field is the velocity field, and the displacement is
secundary.


\section{Notation}
\label{sec:notation}

The velocity field is represented in many ways. Here,
we use bold letters for the whole vector, and subindices
for its components, Cartesian or otherwise:
\begin{align*}
  \bfu = ( u_x, u_y, u_z) \qquad \text{Cartesian} \\
  \bfu = ( u_r, u_\theta, u_z) \qquad \text{Polar, cylindrical} \\
  \bfu = ( u_r, u_\theta, u_\phi) \qquad \text{Polar, spherical} .
\end{align*}



  This naming convention is the most common in physics, and is
  pecified by ISO standard 80000-2:2009, and earlier in ISO 31-11
  (1992). However, the name of the two angles is often reversed,
  specially in mathematics. The advantage of the latter is that
  $\theta$ then retains the same name as in 2D polar coordinates ---
  even though some mathematicians consistently use $\phi$ for the 2D
  polar angle

\section{Style}

In these notes, concepts are introduced \emph{at needed}. This is at
variance with other texts, in which there may be an introduction to
the general physics of fluids, another for mathematical analysis, and
so on. So, for example, the concept of streamlines is not introduced
until the potential flow past a cylinder is discussed. In fact, only
the simple 2D case is introduced --- the more complicated axisymmetric
situation is later described, for the flow past a sphere.

The few starred sections refer to ``advanced'' material, some of which
may be the authors' own calculations (therefore, to be read with
care).
